---
title: 'Project 2'
author: "JL"
date: "02/09/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

########################### <- summary of findings here

```{r, echo = FALSE}
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(stringi)
```

# Loading and Processing the Raw Data

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

```{r, cache=TRUE}
# Download data
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "rawdata.csv.bz2")

# Download documentation for data
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf", "rawdata_documentation.pdf", mode = "wb")

# import data into dataframe
raw <- read.csv("rawdata.csv.bz2")
```

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

## The Data

The dataframe `raw` has 37 columns and 900,000+ observations across the united states. Here is a brief overview.
```{r}
str(raw)
```

Information that we will explore are

- Location: STATE_, COUNTY, COUNTYNAME, STATE
- Time: BGN_DATE, BGN_TIME
- Event Information: EVTYPE, LENGTH, WIDTH, F, REMARKS
- Damage: FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP

For LENGTH and WIDTH, this is the path length (in miles and tenths of miles) and maximum path width (in yards) for all tornadoes.

For F, The "Saffir-Simpson Hurricane and Tropical Cyclone Scale" is used

1. Windspeed 64-82 kts (74-95 mph), storm tide: 4-5 FT, Damage: Minor  
2. Windspeed 83-95 kts (96-110 mph), storm tide: 6-8 FT, Damage: Moderate  
3. Windspeed 96-113 kts (111-130 mph), storm tide: 9-12 FT, Damage: Major  
4. Windspeed 114-135 kts (131-155 mph), storm tide: 13-18 FT, Damage: Severe  
5. Windspeed >135 kts (>155 mph), storm tide: >18 FT, Damage: Catastrophic

For REMARKS, this is a description of the event.

For PROPDMGEXP and CROPDMGEXP, characters are used to signify cost of damage and include “K” for thousands, “M” for millions, and “B” for billions.

## Processing

To process the data for analysis, we will select only the variables that we have outlined above. Further, we will rename these variables to more accessible versions and ensure they are cast into the correct type for calculations. 

Secondly, we will create a dataframe for locations, containing both state and county ids with their respective state and county names, for reference.

```{r, cache=TRUE}
# collect location IDs for reference and remove duplicate rows
counties <- select(raw, "STATE__", "STATE", "COUNTY", "COUNTYNAME") %>%
  distinct(.keep_all = TRUE)

colnames(counties) <- c("state_id", "state_name", "county_id", "county_name")

states <- select(raw, "STATE__", "STATE") %>%
  distinct(.keep_all = TRUE)

colnames(states) <- c("state_id", "state_name")

# states contains duplicates, over separate IDs.
# Manually remove against a list of abbreviations
states <- states[-c(79, 59, 63, 57, 56, 55, 95, 94, 93, 65, 70,
                    60, 68, 61,  66, 73, 71, 72, 76, 74, 62,
                    67, 78, 77, 51, 69, 75, 53, 64, 58, 20, 8), ]


# trim raw to chosen variables
clean <- select(raw, c("STATE__", "COUNTY", "BGN_DATE", "BGN_TIME", "EVTYPE", "LENGTH", "WIDTH", "F", "REMARKS", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP"))

# rename columns
colnames(clean) <- c("state_id", "county_id", "start_date", "start_time", "event", "length", "width", "F", "remarks", "fatalities", "injuries", "property_damage", "property_damage_cat", "crop_damage", "crop_damage_cat")

# cast raw into correct types and formats
clean$state_id <- as.integer(clean$state_id)
clean$county_id <- as.integer(clean$county_id)

clean$event <- stri_trans_totitle(as.factor(clean$event))

clean$F <- as.factor(clean$F)

clean$start_date <- as.Date(as.character(
  strptime(clean$start_date, format = "%m/%d/%Y")))

# We want to include damage_category into damage counts.
# first swap K, M, B with 1e3, 1e6 and 1e9 respectively

clean$property_damage_cat[clean$property_damage_cat == ""] <- 1
clean$property_damage_cat[clean$property_damage_cat == "K"] <- 1e3
clean$property_damage_cat[clean$property_damage_cat == "M"] <- 1e6
clean$property_damage_cat[clean$property_damage_cat == "B"] <- 1e9

clean$crop_damage_cat[clean$crop_damage_cat == ""] <- 1
clean$crop_damage_cat[clean$crop_damage_cat == "K"] <- 1e3
clean$crop_damage_cat[clean$crop_damage_cat == "M"] <- 1e6
clean$crop_damage_cat[clean$crop_damage_cat == "B"] <- 1e9

# now multiply to get correct damage values

clean$property_damage <- as.numeric(clean$property_damage) * as.numeric(clean$property_damage_cat)

clean$crop_damage <- as.numeric(clean$crop_damage) * as.numeric(clean$crop_damage_cat)
```

# Results

## Most harmful events to people's health

This first question we want to investigate is the following

- Across the United States, which types of events are most harmful with respect to population health?

There are a few ways to interpret this. Direct harm, is easily fatalities and injuries, however indirectly we could have the economic impact such as property damage, or crop damage. So we will limit this to the direct harm.

We will group by event and summarise the data

```{r}
 direct_harm <- clean %>%
  group_by(event) %>%
  summarise(number_of_events = n(),
            total_fatalities = sum(fatalities),
            total_injuries = sum(injuries),
            total_sum = sum(fatalities) + sum(injuries)) %>%
  arrange(desc(total_sum))

head(direct_harm)
```

From the summary above, we can see that Tornados have the most fatalities and injuries associated with them. We can also see that the most frequent event is Thunderstorm Wind (Tstm Wind). 

To build a visual, we will adjust the dataframe. Trim to just the top 20 contributing events, sorted by sum of fatalities and injuries. Then collapse fatalities and injuries into one column, but associating a label with each, for ggplot to interpret.

```{r}
# trim a copy of dataframe to top20, select key columns
dh_short <- direct_harm[1:20, ]
dh_short2 <- dh_short

# rename two columns to merge on
colnames(dh_short)[3] <- "Incidents"
colnames(dh_short2)[4] <- "Incidents"

# select only the columns we need for a merge
dh_short <- select(dh_short, c("event", "total_sum", "Incidents"))
dh_short2 <- select(dh_short2, c("event", "total_sum", "Incidents"))

# create a label to split data on plotting
dh_short$label <- "Fatalities"
dh_short2$label <- "Injuries"

# merge dataframes, remove the temprorary second
dh_short <- rbind(dh_short, dh_short2)

rm(dh_short2)
rm(direct_harm)
```

Now we have the data we need for plotting an informational visual in a way that allows us to create it easily.

```{r fig.height=6, fig.width=7}
# plot for the top 20 by total of fatalities and injuries
ggplot(dh_short, aes(x = Incidents,
                 y = reorder(event, total_sum),
                 colour = label),
                 size = 3) + 
  
  theme_economist() + 
  
  scale_colour_stata() +
  
  geom_point() +
  
  labs(x = "Sum of Incidents",
       y = NULL,
       colour = NULL,
       title = "Total Harm caused by Weather Events in\nThe United States",
       subtitle = "Source: National Climatic Data Center Storm Events 1950 - 2011",
       caption = "*Damages to Property are estimates\n1950-54 only tornado events recorded\n 1955-92 only tornado, thunderstorm wind and hail events keyed to digital\n1993-95 only tornado, thunderstorm wind and hail events keyed to digital") +
  
  guides(fill = guide_legend(title = NULL)) +
  
  theme(
    legend.position = "bottom") + 
  
  scale_x_log10(breaks = c(1e1, 1e2, 1e3, 1e4, 1e5),
                labels = c("10", "100", "1,000", "10,000", "100,000"))
```

As you can see from the graph above, for every event, the number of injuries surpasses the number of fatalities. Which is to expected. The exception is Rip Current, which understandably is very dangerous once one has been dragged into one. The most destructive event by far in the U.S. is Tornado, with both regards to total fatalities and total injuries. The latter by a stunning degree. Tornadoes are highly destructive events and they frequent the country.

## Most harmful events with respect to the economy

This second question we want to investigate is the following

- Across the United States, which types of events have the greatest economic consequences?

In this regard, we will investigate the effects of these weather events on damage to Property and damage to Crops. To do so we will group by events and take the sum of property damage and the sum of crop damage, then sort by the total of the two. 

```{r}
economic_harm <- clean %>%
  group_by(event) %>%
  summarise(number_of_events = n(),
            property_damage = sum(property_damage),
            crop_damage = sum(crop_damage),
            total_damage = sum(property_damage) + sum(crop_damage)) %>%
  arrange(desc(total_damage))

head(economic_harm)
```

From the summary above we can see that Flood causes the most expensive damage, over twice the next event, Hurricane/Typhoon. The event most damaging to crops is Drought. Especially noteworthy as it causes relatively low property damage. Floods are second in crop damage, but as already mentioned, deal significant property damage also.

To build a visual, we will adjust the dataframe. Trim to just the top 20 contributing events, sorted by sum of damages. Then collapse property damages and crop damages into one column, but associating a label with each, for ggplot to interpret.

```{r}
# trim a copy of dataframe to top20, select key columns
eco_short <- economic_harm[1:20, ]
eco_short2 <- eco_short

# rename two columns to merge on
colnames(eco_short)[3] <- "Incidents"
colnames(eco_short2)[4] <- "Incidents"

# select only the columns we need for a merge
eco_short <- select(eco_short, c("event", "total_damage", "Incidents"))
eco_short2 <- select(eco_short2, c("event", "total_damage", "Incidents"))

# create a label to split data on plotting
eco_short$label <- "Property Damage"
eco_short2$label <- "Crop Damage"

# merge dataframes, remove the temprorary second
eco_short <- rbind(eco_short, eco_short2)

# rescale damage to thousands of dollars
eco_short$total_damage <- eco_short$total_damage

rm(eco_short2)
rm(economic_harm)
```

Now we have the data we need for plotting an informational visual in a way that allows us to create it easily.

```{r, fig.height=6, fig.width=7}
# plot for the top 20 by total of fatalities and injuries
ggplot(eco_short, aes(x = Incidents,
                 y = reorder(event, total_damage),
                 colour = label),
                 size = 3) + 
  
  theme_economist() + 
  
  scale_colour_stata() +
  
  geom_point() +
  
  labs(x = "Damage in $100,000s",
       y = NULL,
       colour = NULL,
       title = "Total Damage caused by Weather Events\nin the United States",
       subtitle = "Source: National Climatic Data Center Storm Events 1950 - 2011",
       caption = "*Damages to Property are estimates\n1950-54 only tornado events recorded\n 1955-92 only tornado, thunderstorm wind and hail events keyed to digital\n1993-95 only tornado, thunderstorm wind and hail events keyed to digital") +
  
  guides(fill = guide_legend(title = NULL)) +
  
  theme(legend.position = "bottom",
        panel.grid.major.x = element_line(colour = "white",
                                      size = 1,
                                      linetype = "dotted")) +
  
  scale_x_log10(breaks = c(1e5, 1e7, 1e9, 1e11),
                labels = c("1", "100", "10,000", "1,000,000"))

```

ANALYSIS / SUMMARY OF VISUAL

## State that suffers the most

```{r}
state_harm <- clean %>%
  group_by(state_id) %>%
  summarise(number_of_events = n(),
            damage = (sum(property_damage, na.rm=TRUE) + sum(crop_damage, na.rm=TRUE))/1e4,
            fatalities = sum(fatalities),
            injuries = sum(injuries)) %>%
  arrange(desc(number_of_events))

state_harm <- merge(state_harm, states)


# copies of dataframe to merge
state_n <- state_harm
state_p <- state_harm
state_f <- state_harm
state_i <- state_harm

# rename two columns to merge on
state_n$values <- state_n$number_of_events
colnames(state_p)[3] <- "values"
colnames(state_f)[4] <- "values"
colnames(state_i)[5] <- "values"

# create a label to split data on plotting
state_n$label <- "Number of Events"
state_p$label <- "Damages $10,000s"
state_f$label <- "Fatalities"
state_i$label <- "Injuries"

# select only the columns we need for a merge
state_n <- state_n %>% select(c("state_name", "values", "label", "number_of_events"))
state_p <- state_p %>% select(c("state_name", "values", "label", "number_of_events"))
state_f <- state_f %>% select(c("state_name", "values", "label", "number_of_events"))
state_i <- state_i %>% select(c("state_name", "values", "label", "number_of_events"))


# merge dataframes, remove the temprorary dataframes
state_n <- rbind(state_n, state_p)
state_n <- rbind(state_n, state_f)
state_n <- rbind(state_n, state_i)

state_harm <- state_n

# sorting
arrange(state_harm, desc(number_of_events))

# remove temporary dataframes
rm(state_n, state_p, state_f, state_i)
```

```{r, fig.height=14, fig.width=7}
# plot for the top 20 by total of fatalities and injuries
ggplot(state_harm, aes(x = values,
                 y = reorder(state_name, number_of_events),
                 colour = label),
                 size = 3, shape = 22) + 
  
  theme_economist() + 
  
  scale_colour_stata() +
  
  geom_point() +
  
  labs(x = NULL,
       y = NULL,
       colour = NULL,
       title = "Destructive Weather Events in the United States",
       subtitle = "Source: National Climatic Data Center Storm Events 1950 - 2011",
       caption = "*Damages are estimates\n1950-54 only tornado events recorded\n 1955-92 only tornado, thunderstorm wind and hail events keyed to digital\n1993-95 only tornado, thunderstorm wind and hail events keyed to digital") +
  
  guides(fill = guide_legend(title = NULL)) + 
  
  scale_fill_discrete(labels = c("Fatalities",
                                 "Injuries",
                                 "Events",
                                 "Damages $10,000s")) +
  
  theme(
    legend.position = "bottom",
    panel.grid.major.x = element_line(colour = "white",
                                      size = 1,
                                      linetype = "dotted")) + 
  
  scale_x_log10(breaks = c(1e1, 1e2, 1e3, 1e4, 1e5, 1e6, 1e7),
                labels = c("10", "100", "1,000","10,000", "100,000", "1,000,000", "10,000,000"))

```